---
title: "Storm's Health and Economic Impacts"
output: html_document
---

# Loading and Processing Raw Data
This data is coming from U.S. National Oceanic and Atmospheric Adminstraion's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
The data file (repdata-data-StormData.csv) includes the storm data from 1950 to November 2011. The documentation about the data can be found at:

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) 
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r cache=TRUE}
storm.data <- read.csv("repdata-data-StormData.csv", stringsAsFactors = FALSE)
str(storm.data)
```

We are going to look at health and economic impacts of each events arcoss US. So we will need only a few columns from the data -  FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. We will also select Date information - we chose BGN_DATE - along with event type - EVTYPE.

Also column types are modified. We created a date column, the info of which is coming from BGN_DATE column. From this date colume, year column is created. Also EVTTYPE column becomes factor.

```{r cache=TRUE}
columns <- c(2,8,23:28)
storm <- storm.data[,columns]
str(storm)

storm$date <- as.Date(storm$BGN_DATE, "%m/%d/%Y %H:%M:%S")
storm$year <- strftime(storm$date, "%Y")
storm$EVTYPE <- as.factor(storm$EVTYPE)
```

Make Property Damage (PROPDMG) and Crop Damage (CROPDMG) columns as full number, using PROPDMGEXP and CROPDMGEXP ("K" for thousand, "M" for million, "B" for billion, all others are ignored).
Then select only necessary columns.

```{r cache=TRUE}
storm <- transform(storm, PROPDMG = ifelse(PROPDMGEXP == "K", PROPDMG * 1E3, 
                                           ifelse(PROPDMGEXP == "M", PROPDMG * 1E6, 
                                                  ifelse(PROPDMGEXP == "B", 
                                                         PROPDMG * 1E9, PROPDMG))))
storm <- transform(storm, CROPDMG = ifelse(CROPDMGEXP == "K", CROPDMG * 1E3, 
                                           ifelse(CROPDMGEXP == "M", CROPDMG * 1E6, 
                                                  ifelse(CROPDMGEXP == "B", 
                                                         CROPDMG * 1E9, CROPDMG))))
storm <- subset(storm, select = c("date", "year", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG"))
storm$total.dmg <- storm$PROPDMG + storm$CROPDMG
str(storm)
```

Load dplyr library

```{r message=FALSE}
library(dplyr)
```

Group storm data by year and event type and compute the total dmage by the group
```{r cache=TRUE}
storm.by.year.evt <- group_by(storm, year, EVTYPE)
storm.by.year.evt
```

# 10 Most Economically Damaging Events

We compute total damage for each event at each year.

```{r cache=TRUE}
sum.dmg<- summarise(storm.by.year.evt, sum.dmg = sum(total.dmg))
head(sum.dmg)
dim(sum.dmg)
length(unique(sum.dmg$EVTYPE))
```

Dimension of sum.dmg shows that there are 2326 event and year combinations over 985 different event types.  We will look at top 10 economically most sever events in terms of average damage an event cause through all the years. 

We compute mean damage across year by event type. For this, first we group sum.dmg by event type. Then we compute mean damage for each event using summarise. Next we order the output, mean.dmg.by.evt, descendingly by mean value.


```{r cache=TRUE}
sum.dmg.by.evt <- group_by(sum.dmg, EVTYPE)
mean.dmg.by.evt <- summarise(sum.dmg.by.evt, dmg = mean(sum.dmg))
mean.dmg.by.evt <- mean.dmg.by.evt[order(mean.dmg.by.evt$dmg, decreasing = TRUE),]
mean.dmg.by.evt
```

Load ggplot2 library for plotting.
Take only top 10 most severe damage in its cost. Plot them in bar graph

```{r message=FALSE}
library(ggplot2)
```

```{r}
mean.dmg.top <- mean.dmg.by.evt[1:10,]
mean.dmg.top$dmg <- mean.dmg.top$dmg / 1e6
g <- ggplot(mean.dmg.top, aes(x=EVTYPE, y=dmg))
g + geom_bar(stat="identity", fill="green") +  
    theme(axis.text.x = element_text(angle = 90, hjust=1)) +
    geom_text(aes(label=paste(sprintf( "%1.0f", dmg), "M" ), y=dmg + 700, colour = "pink" ),
              size = 3, show.legend = FALSE) +
    xlab("Event Type") +ylab("Cost of Damage (Million)")
```

### Result
According to above bar chart, hurricane/typhoon caused, on average, the most economic damages each year. There are hurricane and hurricane opal events too, 8th and third highest respectively. These two can be considered as one of hurricane/typhoon events too.

# Injuries and Fatalities caused by Storm

```{r cache=TRUE}
sum.inj.fat <- summarise(storm.by.year.evt, total.injuries=sum(INJURIES), total.fatalities = sum(FATALITIES))
sum.inj.fat
```

Group the above by event type then calculate year average count of injuries and fatalities for each event

```{r cache=TRUE}
inj.fat.by.event <- group_by(sum.inj.fat, EVTYPE)
inj.fat.by.event
mean.inj.fat <- summarise(inj.fat.by.event, mean.inj = mean(total.injuries), 
                                   mean.fat = mean(total.fatalities))
mean.inj.fat.sort.by.fat <- mean.inj.fat[order(mean.inj.fat$mean.fat, decreasing = TRUE),]
mean.inj.fat.sort.by.inj <- mean.inj.fat[order(mean.inj.fat$mean.inj, decreasing = TRUE),]
 
mean.inj.fat.sort.by.inj
mean.inj.fat.sort.by.fat

```
We will extract top 10 events that cause most injuries and fatalities and create a union of those events.

```{r cache=TRUE}
top.inj <- mean.inj.fat.sort.by.inj[1:10,]$EVTYPE
top.fat <- mean.inj.fat.sort.by.fat[1:10,]$EVTYPE
top.inj.fat.evt <- union(top.inj, top.fat)
top.inj.fat.evt
```
We got 15 events from the union. We make a data frame with average count from the events.

```{r cache=TRUE}
top.inj.fat <- mean.inj.fat[mean.inj.fat$EVTYPE %in% top.inj.fat.evt,]
top.inj.fat
```

We make top.inj.fat to a tidy dataset using tidyr package. plyr package will be later used for stacked bar chart.
```{r message=FALSE}
library(tidyr)
library(plyr)
```
```{r cache=TRUE}
top.evt <- gather(top.inj.fat, "type", "count", 2:3)
top.evt$type <- as.factor(top.evt$type)
top.evt
```
```{r cache=TRUE}
g <- ggplot(top.evt, aes(EVTYPE,count,fill=type))
g + theme(axis.text.x = element_text(angle = 90, hjust=1)) +
    geom_bar(stat="identity", position="dodge") +
    geom_text(aes(label=sprintf( "%1.0f", count),y = count + 50),position=position_dodge(width=1), size=2.5) +
    scale_fill_discrete(name="", breaks=c("mean.fat", "mean.inj"), labels=c("fatalities","injuries")) +
    xlab("Event") + ylab("Mean Count")
```

```{r cache=TRUE}
top.evt <- ddply(top.evt, "EVTYPE", transform, label_height = cumsum(count) - count / 2 )
```

```{r cache=TRUE}
g <- ggplot(top.evt, aes(EVTYPE,count,fill=type))
g + theme(axis.text.x = element_text(angle = 90, hjust=1)) +
    geom_bar(stat="identity", position="stack") +
    geom_text(aes(label=sprintf( "%1.0f", count), y = label_height), size=2.2) +
    scale_fill_discrete(name="", breaks=c("mean.fat", "mean.inj"), labels=c("fatalities","injuries")) +
    xlab("Event") + ylab("Mean Count")
```


# Information on Current R Environment

Below is our R enviromental information:
```{r}
sessionInfo()
```
