---
title: "Storm's Health and Economic Impacts"
output: html_document
---

# Loading and Processing Raw Data
This data is coming from U.S. National Oceanic and Atmospheric Adminstraion's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
The data file (repdata-data-StormData.csv) includes the storm data from 1950 to November 2011. The documentation about the data can be found at:

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) 
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r cache=TRUE}
storm.data <- read.csv("repdata-data-StormData.csv", stringsAsFactors = FALSE)
str(storm.data)
```

We are going to look at health and economic impacts of each events arcoss US. So we will need only a few columns from the data -  FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. We will also select Date information - we chose BGN_DATE - along with event type - EVTYPE.

Also column types are modified. We created a date column, the info of which is coming from BGN_DATE column. Also EVTTYPE column becomes factor.

```{r cache=TRUE}
columns <- c(2,8,23:28)
storm <- storm.data[,columns]
str(storm)

storm$date <- as.Date(storm$BGN_DATE, "%m/%d/%Y %H:%M:%S")
storm$EVTYPE <- as.factor(storm$EVTYPE)
```

Make Property Damage (PROPDMG) and Crop Damage (CROPDMG) columns as full number, using PROPDMGEXP and CROPDMGEXP ("K" for thousand, "M" for million, "B" for billion, all others are ignored).
Then select only necessary columns.

```{r cache=TRUE}
storm <- transform(storm, PROPDMG = ifelse(PROPDMGEXP == "K", PROPDMG * 1E3, 
                                           ifelse(PROPDMGEXP == "M", PROPDMG * 1E6, 
                                                  ifelse(PROPDMGEXP == "B", 
                                                         PROPDMG * 1E9, PROPDMG))))
storm <- transform(storm, CROPDMG = ifelse(CROPDMGEXP == "K", CROPDMG * 1E3, 
                                           ifelse(CROPDMGEXP == "M", CROPDMG * 1E6, 
                                                  ifelse(CROPDMGEXP == "B", 
                                                         CROPDMG * 1E9, CROPDMG))))
storm <- subset(storm, select = c("date", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG"))
str(storm)
```
